\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-02-19 Writeup/kohler_clarke_2021.tex
reformatting latex

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission
% \usepackage{graphicx}

\title{The human visual system preserves the hierarchy of 2-dimensional pattern regularity}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a, b, 1]{Peter J. Kohler}
\author[c]{Alasdair D. F. Clarke}

\affil[a]{York University, Department of Psychology, Toronto, ON M3J 1P3, Canada}
\affil[b]{Centre for Vision Research, York University, Toronto, ON, M3J 1P3, Canada}
\affil[c]{Stanford University, Department of Psychology, Stanford, CA 94305, United States}
\affil[d]{University of Essex, Department of Psychology, Colchester, UK, CO4 3SQ}

% Please give the surname of the lead author for the running footer
\leadauthor{Kohler}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Wallpaper groups were discovered in the mid-19th century, and the 17 groups constitute the complete set of possible ways of regularly tiling the 2-D plane. In recent years wallpaper groups have found use in the vision science community, as an ideal stimulus set for studying the perception of symmetries in textures. Here we present brain imaging and psychophysical data using the complete set of wallpaper groups and show that the hierarchical organization among wallpaper groups is reflected both in representations in visual cortex and performance on a symmetry detection task. This shows that the visual system is highly sensitive to regularities in textures, and suggest that symmetries may play an important role in texture perception.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{PJK and ADFC designed the study, PJK collected EEG data, ADFC collected psychophysical data, PJK and ADFC wrote the paper.}
\authordeclaration{The authors have no conflicts of interests to declare}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: \href{mailto:pjkohler@syorku.ca}{pjkohler@yorku.ca}}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}

\begin{abstract}Symmetries are present at many scales in images of natural scenes. A large body of literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The goal of the current study is to provide a complete description of responses to symmetry in the human visual system, by collecting both brain and behavioral data using all 17 groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: >0.95 for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: >0.95 for 49 of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

@@

\maketitle
 \thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[h!tb]
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 43/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and 49/64 = 78\% of the comparisons for the primary SSVEP and threshold duration datasets, repectively. This compares to the 32/64= 50\% and 22/64 = 35\% for the SSVEP control datasets.}
\label{fig:durations_rotations}
\end{figure}

% \section*{Guide to using this template on Overleaf}

% Please note that whilst this template provides a preview of the typeset manuscript for submission, to help in this preparation, it will not necessarily be the final publication layout. For more detailed information please see the \href{http://www.pnas.org/site/authors/format.xhtml}{PNAS Information for Authors}.

% If you have a question while using this template on Overleaf, please use the help menu (``?'') on the top bar to search for \href{https://www.overleaf.com/help}{help and tutorials}. You can also \href{https://www.overleaf.com/contact}{contact the Overleaf support team} at any time with specific questions about your manuscript or feedback on the template.

% \subsection*{Author Affiliations}

% Include department, institution, and complete address, with the ZIP/postal code, for each author. Use lower case letters to match authors with institutions, as shown in the example. Authors with an ORCID ID may supply this information at submission.

% \subsection*{Submitting Manuscripts}

% All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}. If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.

% \subsection*{Format}

% Many authors find it useful to organize their manuscripts with the following order of sections;  Title, Author Affiliation, Keywords, Abstract, Significance Statement, Results, Discussion, Materials and methods, Acknowledgments, and References. Other orders and headings are permitted.

% \subsection*{Manuscript Length}

% PNAS generally uses a two-column format averaging 67 characters, including spaces, per line. The maximum length of a Direct Submission research article is six pages and a Direct Submission Plus research article is ten pages including all text, spaces, and the number of characters displaced by figures, tables, and equations.  When submitting tables, figures, and/or equations in addition to text, keep the text for your manuscript under 39,000 characters (including spaces) for Direct Submissions and 72,000 characters (including spaces) for Direct Submission Plus.

% \subsection*{Data Archival}

% PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the Information for Authors, is acceptable.

% % \begin{SCfigure*}[\sidecaptionrelwidth][t]
% % \centering
% % % \includegraphics[width=11.4cm,height=11.4cm]{frog}
% % \caption{This caption would be placed at the side of the figure, rather than below it.}\label{fig:side}
% % \end{SCfigure*}

% \subsection*{Digital Figures}

% Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for figures that will appear in the main text, and images must be final size. Authors may submit U3D or PRC files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.  Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

% Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

% Figure \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide. Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side captions.

% \subsection*{Tables}
% In addition to including your tables within this manuscript file, PNAS requires that each table be uploaded to the submission separately as a “Table” file.  Please ensure that each table .tex file contains a preamble, the \verb|\begin{document}| command, and the \verb|\end{document}| command. This is necessary so that the submission system can convert each file to PDF.

% \subsection*{Single column equations}

% Authors may use 1- or 2-column equations in their article, according to their preference.

% To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.

% Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.

% \begin{figure*}[bt!]
% \begin{align*}
% (x+y)^3&=(x+y)(x+y)^2\\
%        &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
%        &=x^3+3x^2y+3xy^3+x^3. 
% \end{align*}
% \end{figure*}

% \begin{table}%[tbhp]
% \centering
% \caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
% \begin{tabular}{lrrr}
% Species & CBS & CV & G3 \\
% \midrule
% 1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
% 2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
% 3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
% \bottomrule
% \end{tabular}

% \addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
% \end{table}

% \subsection*{Supporting Information (SI)}

% Authors should submit SI as a single separate PDF file, combining all text, figures, tables, movie legends, and SI references.  PNAS will publish SI uncomposed, as the authors have provided it.  Additional details can be found here: \href{http://www.pnas.org/page/authors/journal-policies}{policy on SI}.  For SI formatting instructions click \href{https://www.pnascentral.org/cgi-bin/main.plex?form_type=display_auth_si_instructions}{here}.  The PNAS Overleaf SI template can be found \href{https://www.overleaf.com/latex/templates/pnas-template-for-supplementary-information/wqfsfqwyjtsd}{here}.  Refer to the SI Appendix in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc.

% Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.

% \subsubsection*{SI Datasets}

% Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.

% \matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

@@

%\section*{snippets}
%Specifically, the amplitudes of symmetry-specific responses in individual participants ($n=25$) preserve these relationships at an above-chance level in 88.3\% (53 out of 60) of cases.

%we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. 
%Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of \textit{P1} exemplars, that were matched in terms of their Fourier power spectrum.

%Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1.

\showmatmethods{} % Display the Materials and Methods section

\acknow{PJK acknowledges funding support from the Canada First Research Excellence Fund and the National Sciences and Engineering Research Council of Canada.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliographystyle{pnas-new} 
\bibliography{literature}

>>>>

\documentclass[11pt, twoside]{article}
\usepackage{boisik}
\usepackage[OT1]{fontenc}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{geometry}
 \geometry{a4paper, total={170mm,257mm}, left=20mm, top=20mm}

\title{\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity}
\author[1,2,3]{Peter J. Kohler}
\author[4]{Alasdair D. F. Clarke}

\affil[1]{\small York University, Department of Psychology, Toronto, ON M3J 1P3, Canada}
\affil[2]{\small Centre for Vision Research, York University, Toronto, ON, M3J 1P3, Canada}
\affil[3]{\small Stanford University, Department of Psychology, Stanford, CA 94305, United States}
\affil[4]{\small University of Essex, Department of Psychology, Colchester, UK, CO4 3SQ}

\date{}

@@

\maketitle

\begin{abstract}
Symmetries are present at many scales in images of natural scenes. A large body of literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The goal of the current study is to provide a complete description of responses to symmetry in the human visual system, by collecting both brain and behavioral data using all 17 groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: $>0.95$ for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: $>0.95$ for 49 of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

Symmetries are present at many scales in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.8\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[hptb]
\centering
\includegraphics[width=0.8\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 43/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and 49/64 = 78\% of the comparisons for the primary SSVEP and threshold duration datasets, repectively. This compares to the 32/64= 50\% and 22/64 = 35\% for the SSVEP control datasets.}
\label{fig:durations_rotations}
\end{figure}

@@

% Bibliography
\begin{multicols}{2}
\small
\bibliographystyle{plainnat} 
\bibliography{literature}
\end{multicols}
